The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2020-09-11 Writeup/manuscript.tex
edits to the intro

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work has focused on four of the wallpaper groups, and used functional MRI to show that rotation symmetries within wallpapers are represented parametrically in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust in electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend on this work by collecting SSVEPs and psychophysical data from human participants viewing the complete set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus, with the goal of providing a more complete picture of how wallpaper groups are represented in the human visual system.

@@

The two datasets presented here puts on in the position of being able to assess the extent to which both behavior and brain responses follow that hierachy. The results show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds preserving these relationships at a level that is far beyond chance. Visual cortex thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory. Given that most participants had no knowledge of group theory, the ordered structure of visual responses to wallpaper groups is likely learned implicitly from regularities in the visual environment.

@@

We now wanted to quantify the degree to which our two measurements were consistent with the subgroup relationships. We hypothesized that more complex groups would (a) produce symmetric-specific SSVEPs with higher amplitudes and (b) shorter display duration thresholds. \textcolor{red}{PERHAPS WE NEED TO JUSTIFY THESE ASSUMPTIONS?} We tested each of these hypotheses using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data, and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions could allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 64 subgroup relationships, supergroups are more likely to produce larger symmetry specific SSVEPs and shorter threshold durations, and in most cases the conditional probability of this happening is extremely high.

>>>>

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

The two datasets presented here puts on in the position of being able to assess the extent to which both behavior and brain responses follow that subgroup hierachy. Based on previous imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \cite{RN1170, RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixation presentation time, discriminability increases with the number of symmetry axes in the pattern \cite{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. Visual cortex thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory. Given that most participants had no knowledge of group theory, the ordered structure of visual responses to wallpaper groups is likely learned implicitly from regularities in the visual environment.

@@

We now wanted to test our two assumptions about how SSVEP amplitudes and duration thresholds would follow subgroup relationships, and thereby quantify the degree to which our two measurements were consistent with group theory. We tested each of the two hypotheses using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data, and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions could allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 64 subgroup relationships, supergroups are more likely to produce larger symmetry specific SSVEPs and shorter threshold durations, and in most cases the conditional probability of this happening is extremely high.
